---
title: "Coalitions and Principal-Agent Dynamics: Delegation and the Advocacy Coalition Framework"
author:
- name: Matthew C. Nowlin
  affiliation: Department of Political Science, College of Charleston
  email: nowlinmc@cofc.edu
  git: https://github.com/mnowlin/ACFandPA
date: October 2014
abstract: The Advocacy Coalition Framework (ACF) posits that policy actors within subsystems aggregate into coalitions based on shared values and coordinate to achieve policy objectives. These subsystem actors include elected officials and bureaucrats, which suggests that members of the bureaucracy are likely to belong to one or more advocacy coalitions. However, a large literature suggests that the relationship between elected officials and the bureaucracy can best be understood within a principal-agent framework. In a principal-agent relationship, elected officials---the principals---delegate some policymaking discretion to bureaucrats---the agents---while exercising mechanisms of control to prevent the agents from "drifting" from the policy objectives of the principal. While the dynamics of political control have been well explored, the implications of those dynamics for the policymaking process is less well developed. Combining insights from principal-agent theory and the Advocacy Coalition Framework (ACF), I explore the nature of principal-agent relationships within and across advocacy coalitions using an in-depth case study of used nuclear fuel (UNF) management. Specifically, I development three theoretical propositions regarding principals and agents as members of advocacy coalitions and examine those propositions with the UNF case study. I conclude with a discussion of the need for future research that addresses the implications of agency theory and the ACF on the policymaking process.
conference: Paper prepared for the Theoretical Nexus of Public Policy and Management workshop at the University of Arizona, October 16th--18th, 2014. 
...

```{r, include=FALSE}
bud <- read.csv("OCRWMbudget.csv")
```

# Introduction

Modern industrial countries, such as the United States, are governed through an administrative state that is an increasing complex web of coalitions and networks that are bound together by shared expertise, interests, and/or values. A key concern about the administrative state is keeping it accountable to elected officials, and by extension, the citizens that it governs. Principal-agent theory has been the major theoretical framework through which the issue of administrative accountability has been understood by scholars. In brief, principal-agent, or agency theory posits that principals (e.g., elected officials) delegate some policymaking authority to an agent (e.g., the bureaucracy) that possess some knowledge or skills that the principal does not. In addition, principals engage in monitoring and enforcement to prevent agents from "drifting" from the goals of the principal [for more see @wood_agency_2010]. The institutional arrangements that are shaped by principal-agent relationships exist within specific policy designs and those policy designs are embedded in a larger structure of governance, where governance is defined as "regimes of laws, rules, judicial decisions, and administrative practices, that constrain, prescribe, and enable the provision of publicly supported goods and services" [@lynn_improving_2001, 7]. 

A separate body of work, developed under the rubric of policy theory [see @sabatier_theories_2014], has put forth multiple theories and frameworks to explain the process of policymaking. While this work has provided multiple insights into policymaking, it has yet to incorporate notions of principal-agent or delegation in a comprehensive way [@lavertu_integrating_2009]. Agency theory has greatly enhanced our understanding of the strategic actions of principals in their attempts to control agents, however these insights have, for the most part, not been applied to our understanding of the policymaking process. There are however several points of overlap between work in policy theory and agency theory that could be used to integrate agency theory into the policy process literature. In particular, the Advocacy Coalition Framework (ACF), a well developed framework for understanding the importance of coalitions and policy learning on policy change and development [@sabatier_policy_1993; @weible_themes_2009; @jenkins-smith_advocacy_2014] could be combined with insights gleaned from agency theory to develop a more complete understanding of the nature of the delegation of policy authority from a principal to an agent. The ACF argues that policy actors, including elected officials (the principals) and bureaucrats (the agents), are members of advocacy coalitions that share policy preferences and coordinate actions to achieve shared policy goals. Assuming that bureaucrats are in these coalitions, how do advocacy coalitions impact the dynamics of principal-agent relationships? In addition, how are advocacy coalitions created and sustained within and across various types of policy subsystems? 

In this manuscript I explore how insights from the ACF and agency theory can shed light on the nature of the relationship between principals and agents within the same coalition and that nature of those relationships across different coalitions. In addition, I examine how these relationships are embedded within policy designs and how they may differ across policy subsystems types. Specifically, I develop three theoretical propositions about the nature of coalitions and delegate. The first propositions states that principals and agents are members of advocacy coalitions. This follows directly from the ACF, yet the implications of this for agency theory have not been explored. The second proposition states that bureaucracies are subject to political control by multiple principals and those principals are aggregated into advocacy coalitions. A point of contention in much of the literature on delegation and political control is the question of which institution, Congress or the president, is the most dominant in terms of controlling the bureaucracy. I posit that this question is best viewed through coalitions, namely which coalition is the most dominant and what potentially influential political actors are members of which coalition. Finally, the third proposition posits that delegation and control mechanisms are embedded within policy designs and that the principals preferred role for the bureaucrat in a subsystem will vary by the nature of the conflict within that subsystem. 

I examine these propositions using the policy issue of nuclear waste management in the US. The management of used nuclear fuel (UNF), the potentially hazardous material that results for the production of nuclear energy, has been a vexing policy problem for several decades and involves scientific and technical debates over the proper and handling of storage and normative debates about how and where to site a potentially hazardous facility [@vandenbosch_nuclear_2007; @walker_road_2009]. These factors make it an ideal case to explore coalitions, delegation, and political control given the expertise-driven nature of the question of UNF management, coupled with the politically-driven process of choosing a potential state or territory to host a UNF facility. 

The first section of the manuscript briefly outlines the major insights of the literature on agency theory and delegation. The next section discusses the ACF and the role of bureaucrats within advocacy coalitions. Following that I discuss the nature of delegation and control and how they are embedded within policy designs and I then lay out in more detail the three theoretical propositions. In the fourth section of the manuscript, I briefly discuss the development of the UNF subsystem and UNF management. Following that I use the case of UNF management to examine the theoretical propositions. Finally, I conclude with a discussion of possible future research using agency theory and the ACF.

# Agency Theory and Delegation

The tradition notion of bureaucratic expertise points to a dichotomy between "politics", where debates involve the normative values that inform policy choices and "administration" where debates involve the best way(s) to implement the political decisions. This implies an "overhead democracy" where bureaucrats are objective experts that use their expertise to provide information to elected leaders and implement their decisions in a manner that displays neutral competence [@kaufman_emerging_1956; @redford_democracy_1969; @meier_politics_2007; @workman_policymaking_2010]. Overhead democracy, understood in this way, implies a system where citizens delegate authority to elected leaders and elected leaders, in turn, delegate to experts within the bureaucracy [@moe_new_1984]. This process is best understood using principal-agent, or more simply agency theory [@wood_agency_2010].  Agency theory posits that _principals_ are reliant on _agents_ to perform tasks that the principal is unwilling or unable to perform, principals therefore delegate some policymaking authority to the agent [@epstein_delegating_1999; @bendor_theories_2001; @lavertu_integrating_2009]. However, bureaucrats may make strategic use their expertise if their policy preferences differ from that of the principal [@mitnick_theory_1975] to gain more resources [@niskanen_bureaucracy_1971], claim policy jurisdiction [@nicholson-crotty_bureaucratic_2005], and/or build decision-making autonomy [@carpenter_forging_2001]. Therefore, elected officials must exercise some level of "control" over bureaucracies to ensure that their preferences remain reasonably well-aligned. As a result of the possible tensions inherent in the principal-agent relationship, a large body of work has been built around understanding the dynamics of delegation and control.

The delegation of policymaking authority from elected officials has been examined in the context of Congressional Dominance [@weingast_bureaucratic_1983; @mccubbins_congressional_1984], where Congress is the primary principal, Presidential Dominance [@moe_presidents_1994], where the president is the key political actor, and a multiple-principal approach [@spiller_politicians_1990; @hammond_who_1996;  @warren_allies_2012], which posits that multiple principals including congress, the president, and the courts delegate authority and/or exercise control over bureaucratic agencies. I largely draw from the multiple-principals approach, however I argue that advocacy coalitions rather than institutional affiliation is the lens through which delegation should be viewed. 

According to a review by @huber_politics_2006, scholars have laid out four arguments about the nature of delegation, irrespective of the political institution (Congress, the president) of the elected official. First, elected officials are more likely to delegate to the bureaucracy when there is greater _policy uncertainty_. Elected officials are concerned about achieving particular outcomes, yet they are uncertain about the connection between a policy and its outcome, therefore they rely on information from experts [@esterling_political_2004; @burstein_interest_2007]. As a result, when policy uncertainty is high elected leaders are more likely to delegate to the bureaucracy [@bawn_political_1995]. Apart from policy uncertainty, the _ally principle_ states that "as the policy preferences of politicians and bureaucrats converge politicians will delegate more discretion to bureaucrats" [@huber_politics_2006, 260]. In contrast, as preferences diverge politicians become less likely to grant significant autonomy to bureaucrats.

Separate from the granting of authority, elected officials create control mechanisms to limit the discretion of bureaucrats. These control mechanisms can be _ex ante_, where they are prior to policy implementation and/or _ex post_, where they following policy implementation [@mccubbins_administrative_1987; @mccubbins_structure_1989]. Often elected officials are indifferent with regard to _ex ante_ or _ex post_ controls, therefore the literature has identified a  _substitution effect_ between the two types of control mechanisms [@bendor_spatial_2004]. The fourth argument presented in the literature is about _political uncertainty_. Political uncertainty deals with the concerns of elected officials that future political choices will alter their current policy choices based on a shift in political preferences. Therefore, elected officials design policies that limit, or "insulate", bureaucratic discretion to protect against future political choices [@de_figueiredo_electoral_2002]. 

In general, politicians have incentives to delegate some policymaking authority to bureaucracies and, all else equal, are more likely to do so when policy uncertainty is high and the bureaucracy is considered an ally--meaning it shares the policy preferences of the principal. However, elected officials also have reasons to be concerned about bureaucratic drift and therefore they implement a serious of controls to reign in bureaucratic behavior. While the literature on political control is well developed, questions still remain about what bureaucrats are likely to do with the discretion they are given and/or what factors would motivate bureaucratic drift [@krause_legislative_2010]. Drawing on the distinctions between the various roles that policy analysts can play in a policy subsystem put forward by @jenkins-smith_democratic_1990, bureaucrats can be thought of as "objective technicians", "issue advocates", or "client advocates." Objective technicians embody the notion of neutral competence, where bureaucrats remain separate from the political process and seek only to use facts and unbiased analysis to determine the best, most efficient, program to implement the goal of the political principal. At the other end of the spectrum are issue advocates, that are driven by normative beliefs about what policy _should_ be and they are active participates in the political process. Finally, client advocates are also active in the political process, however they advocate positions that the client prefers. @jenkins-smith_democratic_1990 also notes that the type of role that the analyst can be inhibited or encourage depending on the context of the policy subsystem. For example, objective technicians are more likely to thrive in environments that are high in analytical tractability, low in conflict, have closed or professional forums, and have low levels of organizational allegiance. Issue advocates tend to thrive in subsystems with low analytical tractability, high conflict, open forums, and low organizational allegiance. Finally, client advocates thrive in the same environments as issue advocates, except that client advocates are more likely when organizational allegiance is high [@jenkins-smith_democratic_1990, 117-118]. The normative expectations of bureaucrats in an overhead democracy are that they would act as objective technicians. However, objective technicians are only likely in a specific type of subsystem [also see @weible_expert-based_2008]. Therefore, it is likely that the roles that bureaucrats adopt are competing, overlapping, and based on the context of the subsystem in which they operate. 

Some of the early work on the behavior of bureaucrats also posited that they exhibit mixed motivations and play multiple possible roles in the policy process. @downs_inside_1967 argued that bureaucrats can be  _self-interested_ as either _climbers_ or _conservers_ or that they combine self-interest with normative values and/or loyalty and act as _zealots_, _advocates_, or _statesmen_.^[Apart from the pure self-interest categories, the categories developed by Downs overlap somewhat with those of Jenkins-Smith, particularly with regard to advocates.] One way that bureaucrats have been argued to be self-interested is that they are budget maximizers and seek to have budgets that are above what is socially optimal [@niskanen_bureaucracy_1971], although the empirical evidence for Niskanen's model is inconclusive at best [@blais_budget_1991]. However, the notion self-interest can be broader than maximizing budgets and could include interest in terms of policy preferences, such as with issue advocates that are motivated by normative beliefs or client advocates that are motivated by loyalty to the client. The policy preferences of the bureaucrat are an important component of agency theory, and some evidence suggests that the preferences of bureaucrats often diverge from that of the elected officials [@aberbach_web_2000]. In addition, @carpenter_forging_2001 argued that bureaucrats act as entrepreneurs, or advocates, for their agencies and/or their policy goals.  This implies bureaucrats that are likely to be _policy-motivated_, therefore less motivated by power or prestige (i.e., self-interest) and less likely to be pure objective technicians. Indeed, @gailmard_slackers_2007 found that the development of policy expertise is more likely when public servants are policy-motivated. They argue that "neutral competence is impossible in our model not because 'neutrality' is impossible, but because only those with a stake policy can be induced to become experts" [@gailmard_slackers_2007, 887]. In other words, bureaucrats self-select to become experts based on policy-motivations. If correct, then the development of expertise and policy preferences are endogenous and the question of what drives the policy preferences of bureaucrats ought to be paramount with regard to questions of political control. In addition, this makes the role of objective technician less likely. The question then is, what drives the policy motivations of bureaucrats? It is likely that the policy preferences and motivations of bureaucratic actors derive, much like other policy actors, from their belief systems as posited by the Advocacy Coalition Framework [@sabatier_policy_1993; @jenkins-smith_advocacy_2014]. 

# The Advocacy Coalition Framework
The Advocacy Coalition Framework (ACF) posits that policymaking occurs within and across semi-autonomous policy subsystems, and these subsystems are the "primary unit of analysis" when examining the policy process [@jenkins-smith_advocacy_2014, 189]. Subsystems are centered around specific policy areas/domains and include interested policy actors that hope to influence policy decisions within that particular domain. The ACF argues that these actors aggregate into one or more advocacy coalitions based on shared beliefs and coordinate actions and resources to achieve shard policy goals. The beliefs that are central to the ACF are understood as a system that has  three-tiers that are hierarchically arranged from the most general (deep core beliefs) to the most specific (secondary beliefs). In brief, the belief system includes deep core beliefs that are fundamental axioms (e.g., political ideology) that span multiple policy areas and guide (or constrain) more domain specific beliefs, termed policy core beliefs. Policy core, or subsystem specific, beliefs are policy area specific beliefs and serve as the organizing beliefs for coalitions within a subsystem. Finally, secondary beliefs are the most specific and include the means that could be used to achieve policy goals. 

The types of policy actors that constitute advocacy coalitions include the tradition "iron triangle" of actors such as members of Congress, interest groups, and bureaucratic agencies, however they also include other actors including state and local government officials, non-profit groups, governmental contractors, scientists, and members of the media. As noted, what binds these actors into coalitions are shared values, beliefs, and preferences regarding policy choices [@weible_comparing_2005; @henry_ideology_2011]. The notion that elected officials--the principals--and bureaucrats--the agents--are members of advocacy coalitions have important implications for agency theory, although these implications have not been fully explored.^[Previous work on principal-agent dynamics and the ACF noted that purposive interest groups, due to their focus on normative and/or ideological values, might be more constrained in the types of policies that they can accept [@jenkins-smith_politics_1993].] Related to this, little work has examined the role of elected officials, particularly visible politicians like members of Congress and the president, in advocacy coalitions. Given that these elected officials can be a vital component of policy development and change, their role, particularly as principals, should be further explored within the ACF. 

While scant research has examined elected officials and politicians as part of advocacy coalitions, several studies have found that bureaucrats are indeed members of advocacy coalitions [@jenkins-smith_explaining_1991; @jenkins-smith_politics_1993; @ellison_advocacy_1998; @sabatier_are_1999; @weible_comparing_2005]. In addition, given their expertise and resources agency officials may be central to coalition success   [@ellison_building_2010]. Much like was argued by @gailmard_slackers_2007,  the notion that bureaucrats are policy-motivated members of advocacy coalitions goes against traditional ideas of neutral competence and agency responsiveness and implies that bureaucrats are policy motivated and value driven just like other policy actors. 

The connection between core values and beliefs and bureaucratic behavior is a consistent theme in public administration scholarship, particularly the research on Public Service Motivation (PSM) [@perry_motivational_1990; @houston_public-service_2000; @perry_bringing_2000; @lee_behavioral_2012], and a number of recent studies have examined ideology and it's impact on the behavior of agency officials and other public managers. For example, more liberal managers tended to view the Program Assessment Rating Tool (PART), developed during the George W. Bush administration, as more onerous [@lavertu_government_2013]. Another study found that as university presidents became more conservative, they became more supportive of performance-based funding for universities [@rabovsky_support_2014]. In addition, a growing body of work is interested in measuring the ideal points along a uni-dimensional scale of ideology of executive agencies to allow comparisons between agencies, the president, and members of Congress [@clinton_expert_2008; @bertelli_lengthened_2011; @clinton_separated_2012]. Finally, using the ACF  @sabatier_are_1999 found that agency officials held similar beliefs as the interest groups within their coalition.

The above research provides some evidence that bureaucrats are driven by deep core beliefs that influence their policy core beliefs.  However, it has been hypothesized that agencies will take a more moderate stance relative to other members of their coalition [@jenkins-smith_advocacy_2014, 195].^[Indeed, this is stated as _Coalition Hypothesis 4_, although as noted by @jenkins-smith_advocacy_2014, it has rarely been tested.] One of the few studies that examined this hypothesis found that executive agencies active within the outer continental shelf oil and gas leasing subsystem tended to take a position between the pro-leasing and anti-leasing coalitions [@jenkins-smith_politics_1993]. In addition, @jenkins-smith_politics_1993 also found that the positions of executive agencies shifted as a result of the election of Ronald Reagan. This moderation and responsiveness to a change in presidents is likely due to the constraints embedded in the principal-agent relationship. As noted by @jenkins-smith_democratic_1990, the context of the policy subsystem can be an important factor in determining the type of role that policy analysts, and by extensions bureaucrats and other experts, play in the policy subsystem. Therefore, that context should be extended to include the governance structures, that include delegation and control mechanisms, within and across policy subsystems. 

# Coalitions, Delegation, and Governance 

The literature on delegation and advocacy coalitions has largely developed in parallel without much interaction, despite the fact that they often address overlapping themes. The delegation literature notes that elected officials grant some policy authority to bureaucrats based on the agencies expertise, while simultaneously exercising control and oversight of the agency. The literature developed using the ACF argues that elected officials and bureaucrats, along with other policy actors, are members of advocacy coalitions that are bound by shared values and beliefs and coordinate to achieve shared policy goals. The implications of principals and agents within the same and/or competing coalitions has not been fully explored. In particular, what are the dynamics of delegation and control across different contexts or subsystem types? In addition, how does having powerful and visible elected officials within a coalition impact coalition behavior and success? 

To address these questions I posit that delegation, control, and oversight  mechanisms are embedded within the policy designs that coalitions enact and that these policy designs form the foundation of _governance_ within the particular policy area, subsystem, or regime. Policy designs are defined as the "content" of policy and include the "rules" (i.e., institutional arraignments) through which actions are required (forbidden), the actors that should undertake those actions, and sanctions for noncompliance [@schneider_policy_1997; @siddiki_dissecting_2011, 81]. In addition, policy designs typically include the policy core beliefs of the dominant coalition [@jenkins-smith_advocacy_2014] and an implementation structure of organizations and agencies charged with implementing the policy [@hall_structures_2000]. Control and oversight mechanisms, often included as part of policy design, provide ways for elected officials, acting as principals, to monitor the implementation of the policy by bureaucrats, acting as agents. Increasingly, these implementation structures have become less hierarchical and more focused on networks of government agencies, contractors, and non-profit organizations [@frederickson_whatever_2005]. Therefore, agency theory should incorporate the various types of organizations and agencies that are charged with implementing public policy and the networks in which they exist. In brief, policy designs with embedded institutional arrangements, policy core beliefs, implementation structures, and oversight mechanisms form the foundation of governance within a particular policy area [@lynn_improving_2001].

As noted, agencies can potentially adopt differ roles--objective technicians, issue advocates, or client advocates--in a subsystem and policy designs can be used to provided incentives for agencies to act in a particular way. In addition, policy designs can include specific _ex ante_ and _ex post_ control mechanisms to influence agency behavior. Coalitions, particularly the political principles within them, are likely to prefer that bureaucrats play certain roles and prefer various levels of delegation and control depending on the nature of conflict within a subsystem [@moe_politics_1995]. As noted, increased political uncertainty makes it more likely that principals will, _ex ante_ as policy is designed, seek to limit bureaucratic discretion in an effort to limit the impact that competing coalitions, present or future, can have on policy implementation and development. Therefore, the type of policy designs are likely a function of the type of subsystem conflict. The next section outlines several theoretical propositions that result from the consideration of the literature on agency theory, the ACF, and governance. 

## Theoretical Propositions 

One of the key insights of the ACF, is that the multiple actors, including elected officials and bureaucrats, within a policy subsystem aggregate into one or more advocacy coalitions. Therefore, the first proposition states that:

\noindent __Proposition 1__: _Principals and agents are in advocacy coalitions_

The notion that principals and agents are members of advocacy coalitions has important implications for theoretical development for both agency theory and the ACF. First, highly visible political principals are important resources for advocacy coalitions. Powerful members of Congress and presidents are often effective advocates and command powerful resources that can be used to achieve coalition goals. The same is also true of bureaucratic agencies, coalitions are able to leverage bureaucratic expertise and resources in the pursuit of policy goals and reduce policy uncertainty [@ellison_building_2010]. 

In addition, using the ACF as a lens to examine agency theory can provide insights to the nature of delegation and discretion. According to the ally principle, politicians are likely to grant more discretion to agencies that share their policy preferences. This insight clearly leads to the expectation that, all else equal, agencies in the same advocacy coalition as the political leaders would receive more discretion. 

Finally, placing bureaucrats within advocacy coalitions ties into a growing literature that addresses the beliefs and values of bureaucrats and the impact of those values on behavior. Depending on their value orientation and the nature of subsystem conflict bureaucrats may act as objective technicians, issue advocates, or client advocates. As noted, mounting evidence suggests that bureaucrats are, like other subsystem actors, policy-motivated and driven by deep core and policy core beliefs that inform their preferred policy choices. However, as discussed more below, often the role a bureaucratic actor takes within a policy subsystem is prescribed within policy designs, therefore bureaucrats are often faced with being policy-motivated and yet constrained by the nature of their prescribed role and the control mechanisms within the governance structure they operate. These cross-cutting and competing influences likely lead to bureaucratic actors being more moderate in their expressed positions, as compared to other members of their advocacy coalitions [@jenkins-smith_advocacy_2014]. 

One of the major controversies within the political control literature is whether Congress or the president should be seen as the dominant political principle. While not fully resolved, many scholars have noted that agencies are controlled by multiple principals that include Congress and the president, but also include the courts and interest groups. Building on the insight of multiple principals, I propose that questions of dominance and political control are best understood through coalitions. 

\noindent __Proposition 2__: _Multiple coalitions instead of multiple principals_

Examining multiple principals through coalitions is directly related to proposition 1. Given that principals and agents are members of advocacy coalitions, it is important to consider which principals and which agents are in which coalition. Specifically, what political actors are in the dominant advocacy coalition and what political actors are in the minority coalition(s). 
The dominant coalition controls the subsystem and likely prefers the status-quo, whereas the minority coalition(s) is trying to influence the subsystem and is likely to prefer policy change. If members of the dominant coalition are in positions of political power, then it is likely that bureaucrats will be given a high degree of discretion, but if elected officials in the minority coalition have positions of power then they are likely to exert political control at both the _ex ante_ and _ex post_ phase. Based on the substitution effect, elected officials are indifferent as to whether political control occurs within the initial design of the policy, _ex ante_ or as the policy is implemented, _ex post_. Both types of control are included as part of the policy design, however they vary as to when they are activated, with _ex ante_ controls taking effect immediately and _ex post_ controls being triggered if needed as the policy is implemented [@mccubbins_congressional_1984]. Therefore, theoretical advancements in understanding the dynamic relationships between principals and agents could be made by utilizing the notion of competing coalitions that contain multiple principals and multiple agents. 

The _ex ante_ granting of discretion and some possible _ex post_ control mechanisms are set in place prior to the implementation of policy. The policy designs that include these features are a result of political conflict, therefore I propose that: 

\noindent __Proposition 3__: _Delegation and control are embedded in policy designs and influenced by the level of conflict in the subsystem_

The types of governance structures that are formed, at least in part, through policy designs can be reflective of the type of coalitions and conflict that exists within the policy subsystem. As noted, the context of a policy subsystem, in terms of it analytical tractability and conflict, can determine the type of role that bureaucrats play in the subsystem [@jenkins-smith_democratic_1990]. One way this can happen in through the process of conflict and comprise in which policy designs are created [@moe_politics_1995; @bertelli_role_2006]. In a policy subsystem with low conflict, policy designs are likely to encourage agencies to act as objective technicians and search for the most effective policies and/or means to implement policies. However, in a highly conflictual subsystem, coalitions would likely want to design policy in such a way as to ensure that bureaucratic agencies are advocates and remain within the coalition. In addition, it is also likely that subsystem type can influence the nature of political control within the subsystem, with more conflict leading to more control and less conflict leading to more discretion. Finally, the preferred role of bureaucrats is likely to vary by coalition type, with the dominant coalition willing to grant agencies more discretion, whereas the minority coalition, likely preferring some policy change, would argue for less discretion in high conflict subsystems. These relationships are depicted in Table \ref{tab:types}. 

\begin{table}[!htp]
\centering
\caption{Preferred Bureaucratic Role and Delegation by Coalition Type and Subsystem Type \label{tab:types}}
\begin{tabular}{l | p{4.5cm} | p{4.5cm} }
\hline
 Coalition Type & Low Conflict Subsystem & High Conflict Subsystem \\
 \hline
 Dominant Coalition & Objective Technician/High Discretion & Client Advocate/High Discretion \\
 Minority Coalition & Client Advocate/High Discretion & Objective Technician/Low Discretion \\
\hline
\end{tabular}
\end{table}

As shown, it is expected that the dominant coalition would prefer to grant agencies more discretion in any case, however the type of role they would prefer the agency to take varies by the level of conflict in the subsystem. Note however, that in no case would any coalition prefer a bureaucrat that is an issue advocate. In subsystems of low conflict the dominant coalition would prefer bureaucracies that are objective technicians and would design policy that gives agencies wide discretion. This is because the dominant coalitions wants the policy that they control to be implemented in the most effective way and therefore grants the expert bureaucracy wide latitude in implementation. In addition, it is likely that the dominant coalition would include the bureaucracy, therefore more discretion is assumed because of the ally principal. Also note that I would expect dominant coalitions to prefer objective technicians with high discretion in subsystems with no opposition or no minority coalition. The minority coalition in low conflict subsystems would prefer agents that act as their advocate and are given wide discretion to do so. Minority coalitions that are seeking to gain influence over the dominant coalition and/or seeking policy change would like expert bureaucrats to seek to advance the goals of the coalition and not be subject to control of the dominant coalition.

In policy subsystems that are high conflict however, the dominant coalition would prefer bureaucrats that act as client advocates. Acting as a client advocate means that the agency leverages its expertise in the pursuit of the goals of the political principals within the dominant coalition. In addition, given that the dominant coalition, by definition, controls the subsystem it is likely that the coalition would, through the design of the implementation structures, seek to ensure that agencies are advocates and within the coalition. Ensuring agencies are in advocacy coalitions helps to guard against the political uncertainty around whether the current dominant coalition will always be the dominant coalition within the subsystem. Assuming that the agency is an advocate, the dominant coalition would prefer more discretion to allow agents to leverage their expertise and to prevent political principals in the minority coalition opportunities to exercise control. The minority coalition in a highly conflictual subsystem would prefer bureaucracies that are objective technicians, yet they would also prefer the ability to have control over the agency to keep them from drifting from the coalition goals. Minority coalitions that are trying to gain a foothold into a subsystem would prefer agencies that produce objective information, which is _more likely to be seen as credible_ by those outside the subsystem. Objective information that undermines the policy core beliefs of the dominant coalition could be a powerful tool for the minority coalition to expand conflict within the subsystem. However when the level of conflict is high, the minority coalition would be prefer that agencies be given less discretion so they could exercise control over the agencies in the dominant coalition and prevent agencies in their own coalition from drifting.

The above discussion detailed three theoretical propositions regarding the nature of delegation and control through the lens of advocacy coalitions. First, it was noted that political principals and bureaucratic agencies are members of advocacy coalitions. Second, it is important to consider the distribution of agencies and elected official within and across different coalitions. Assuming that agencies are subject to political control by multiple political principals, understanding which coalitions the principals and agents are in can provide valuable insights. Finally, it was noted that 
the preferred roles of bureaucrats and the preferred level of delegation vary by coalition types and subsystem conflict. It is expected that dominant coalitions would always prefer to grant more discretion, yet they would prefer objective technicians when conflict is low and client advocates when conflict is high. Minority coalitions however, would prefer client advocates with high discretion with low conflict and objective technicians with low discretion when conflict is high. It should be noted however, that policy designs are often the result of compromise between a dominant and minority coalition. Therefore, the type of delegation and the roles assumed by agencies may not perfectly match the preferences of either the dominant or minority coalitions. Indeed, the networked structure of policy implementation ensures that multiple agencies are involved, which means a mix of agency roles and delegation is more likely. The next section briefly describes the development of used nuclear fuel (UNF) management in the US, the case study I use to examine the nature of coalitions, delegation, and control.

# Case Selection: Used Nuclear Fuel Management

Used nuclear fuel, or nuclear waste, refers to the radioactive material that remains following the production of energy from nuclear fission. Currently this material is stored on-site at various nuclear power plants across the country. During the initial development of civilian nuclear energy, this waste was not considered to be problematic. Indeed, it was largely thought this material could be "reprocessed" as a source of addition fuel and the material that would remain could be easily dealt with  [@walker_road_2009]. However, the nuclear energy subsystem, where debates about UNF occurred, began to undergo profound shifts beginning in the 1960s as the image of nuclear energy shifted from positive to negative and new policy venues were exploited by nuclear energy opponents [@baumgartner_agendas_1993]. These shifts lead to the collapse of the "iron triangle" nuclear energy subsystem and spawned a much more open and contentious subsystem. The development of used nuclear fuel management as its own subsystem followed from this collapse [@walker_road_2009; @nowlin_policy_2014]. Central to the evolution of the UNF subsystem were several key policymakers and policy designs that shaped the type of UNF disposal (a deep geologic repository); the agencies involved with the disposal and regulation of UNF; and the way that potential host sites for disposal were considered and chosen. Throughout most of its evolution, the UNF subsystem had medium to high levels of conflict. 

Once the reprocessing of nuclear material was abandoned in the 1970s, due to concerns over nuclear proliferation, the management of that material became a more pressing issue. The first major policy designed to deal with UNF management, the Nuclear Waste Policy Act of 1982 (NWPA), set-up the process by which the federal government would construct a permanent repository to store UNF. Among its other provisions the NWPA:

- Required the Department of Energy (DOE) to examine five possible locations for a permanent repository site and recommend three to the president by 1985

- Required DOE to perform environmental assessment of each possible site

- Required DOE to apply to the Nuclear Regulatory Commission (NRC) for a license to construct a repository 

- Established the Office of Civilian Radioactive Waste Management within the DOE to oversee the development of the UNF repository

- Established the Nuclear Waste Trust Fund, where fees are collected from the utilities and set aside to pay for the development and operation of the repository  [@cq_almanac_comprehensive_1982; @clary_enactment_1991; @vandenbosch_nuclear_2007].

Following passage of the NWPA, the UNF subsystem became more contentious as states and elected officials became concerned that there state would be chosen to host the repository [@carter_nuclear_1987; @kraft_public_1991]. Indeed, the NWPA only seemed to heighten the involvement of intergovernmental actors (state, local, and tribal officials) in the UNF subsystem as they sought to insure their state or territory was not chosen to host a repository [@nowlin_policy_2014]. The NWPA allowed potential host states a veto over the repository, however that veto could be overridden by Congress. In order to speed development of the repository, Congress, lead by Senator J. Bennett Johnston of Louisiana, in the Nuclear Waste Policy Amendments Act of 1987 (NWPAA) decided that Yucca Mountain in Nevada should be the only site considered to host the UNF repository [@cq_almanac_nevada_1987]. In addition, the NWPAA established the Nuclear Waste Technical Review Board as an independent agency to oversee the DOE's development of UNF storage.

The next major piece of legislation, the Energy Policy Act of 1992 (EnPA), required the National Academy of Sciences to examine radiation disposal and exposure standards which the Environmental Protection Agency (EPA) would then use as the basis of new exposure standards for Yucca Mountain. Following the provisions of the NPWA, DOE recommended Yucca Mountain as the permanent UNF repository to President George W. Bush in 2002 and Bush approved. The state of Nevada veto the approval and that veto was overridden by Congress [@cq_almanac_bush_2002]. The DOE submitted the license to construct the UNF repository at Yucca Mountain to the NRC in the summer 2008, however that license was subsequently withdrawn by the Obama administration in 2010. Following the license withdrawal, President Obama established a blue ribbon commission to examine how UNF storage can be resolved and the commission issued their report in 2012. 

## Used Nuclear Fuel Coalitions

Proposition 1 stated that both principals and agents are in advocacy coalitions. This is evident in the UNF subsystem, where the coalition that was involved with the development of the NWPA included industry groups that were concerned with the amount of UNF that was accumulating at nuclear power plants; scientists and engineers from the DOE that viewed UNF disposal as a technical problem that could be resolved; and some members of Congress, most notably J. Bennett Johnson of Louisiana. The goal of this coalition was to solve the problem of the UNF storage. Those that opposed the NWPA included environmental groups and members of Congress concerned about the siting of the permanent repository in their states or districts [@cq_almanac_comprehensive_1982; @carter_nuclear_1987; @clary_enactment_1991]. Given the rising debates over site selection and concern about delays within the DOE, Congress amended the NWPA to limit site characterization activities to Yucca Mountain in Nevada. This action solidified two coalitions within the UNF subsystem 1) a pro-Yucca Mountain coalition consisting of the nuclear industry, the DOE, and some members of Congress and 2) an anti-Yucca Mountain coalition consisting of the state of Nevada and environmental groups. The pro-Yucca coalition could be considered dominant because they, post-NWPAA, represented the status-quo, held majorities in Congress, and had the DOE working to development Yucca Mountain. The minority anti-Yucca coalition grew more powerful over time, however, and included some high-profile political principals such as President Clinton, who vetoed a bill that would have established Yucca Mountain as a temporary repository in 2000 [@cq_almanac_clinton_2000], and President Obama, who ended consideration of Yucca Mountain as a permanent repository. The most notable member of the anti-Yucca coalition was Senator Harry Reid from Nevada, who became minority leader in 2005 and then majority leader in 2007. 

If coalitions contain both the elected officials as principals and bureaucracies as agents then, according to the ally principle, it is likely that elected officials allow agencies in the same coalition more discretion. This is not as clear in the case of the discretion given to the DOE for the development and characterization of potential host sites for a UNF repository. First, the process laid out in the NWPA was short-circuited by the NWPAA and its designation of Yucca Mountain as the only site to be considered. The NWPAA can certainly be see as limiting the DOE's discretion in selecting the site, but once Yucca Mountain was selected the DOE had rather wide discretion with regard to the processes of site characterization at Yucca Mountain. However, the governance structure that emerged around UNF management contained several control mechanisms including the number of agencies overseeing the DOE and, most notably, the control that Congress exhibited over the DOE's budget for UNF repository development. 

Related to proposition 1, proposition 2 stated that agencies are under the purview of multiple principals and these principals are aggregated into one or more advocacy coalitions. The implication of this is that agencies are subject to the push-and-pull of principals within competing agencies. This was certainly the case in the UNF subsystem given that the DOE, though part of the dominant pro-Yucca coalition, was still subject to oversight by Congress, the president, the courts, and other agencies including the NRC and the EPA. Given the structure of policymaking institutions in the US, the minority coalition can exploit multiple principals across different policy venues. For example, in a case brought by the State of Nevada and environmental groups the D.C. Court of appeals ruled in 2004 that the radiation exposure guidelines for Yucca Mountain set by the EPA did not adequately follow the recommendations of the NAS report, as required by the EnPA of 1992, and must be re-considered [@vandenbosch_nuclear_2007]. By using the courts, the minority coalition was able to limit agency behavior through the use of multiple principals across multiple policy venues. 

As noted, the DOE was in the dominant coalition in support of Yucca Mountain, however the NRC and the EPA seemed to take more moderate stances and also seemed to take the role of objective technician. For instance, in Congressional testimony given in 2002 at a hearing regarding the recommendation of Yucca Mountain by the DOE and President Bush, Greta Joy Dicus from the NRC she stated that:

> If the President's recommendation becomes a final decision, several important steps must be taken before the Commission can decide whether or not to authorize construction of a potential repository at Yucca Mountain. First, DOE must submit a high quality application. Second, our staff at the NRC must conduct an independent safety review and issue a safety evaluation report. And third, we must conduct a full and fair public hearing on the DOE application. Only after these steps are complete will the NRC be in a position to determine based on all of the information before us, whether the DOE's license application complies with our regulations [@dicus_review_2002]. 

This testimony reflects the position of the NRC as the regulator of the DOE and it offers new insight into the arraignment of implementation structures that involve multiple agencies. In addition, it illustrates how the NRC saw its role as one of an objective technician that would judge the legitimacy of the DOE's work on a scientific and technical basis. Previous work has noted that agencies compete for policy influence [@nicholson-crotty_bureaucratic_2005], however in the case of the UNF subsystem both the NRC and the EPA are regulators of the DOE. Therefore, in effect these other agencies become another principal that can 1) limit the discretion of other agents and 2) become tools for principals in competing coalitions to use. 

The third proposition stated that delegation and control are embedded in policy designs and influenced by the level of subsystem conflict. As a result, I posited that the dominant coalition would always prefer to grant discretion to expert agencies, yet prefer client advocates, as opposed to objective technicians, in subsystems with high conflict. The UNF subsystem was, in large part, always conflictual, even as the degree and nature of conflict varied. Therefore, I would predict that the dominant coalition would prefer agencies to be client advocates with a large amount of discretion and design policy accordingly. The implication of this is that the agencies would become part of the dominant coalition. Prior to the NWPA, conflict centered around the choice of disposal type, deep geologic vs. monitored retrievable storage [@vandenbosch_nuclear_2007]. The NWPA gave the DOE--specifically the newly created Office of Civilian Radioactive Waste Management (OCRWM) within DOE--rather high discretion to carry-out site characterization and recommend a site to the president to host a repository. However, instructing the DOE to characterize and recommended a site  _ensured that the DOE would share the goal of the coalition_, thereby making the DOE, in effect, a client advocate and placing them  within the dominant advocacy coalition. This was as expected, the dominant coalition of nuclear energy companies and members of Congress designed a policy that ensured the DOE would leverage its expertise to pursue the goal of siting a repository. In addition, in the NWPA the DOE was given a high amount of discretion in the pursuit of those goals. However, the fact that the DOE mission is one based in energy development, therefore they held the relative expertise is also certainly another reason that the DOE was charged with the development of an UNF repository [@clary_enactment_1991]. Yet, giving the DOE the responsibility of developing a repository changes the nature of their role in the policy subsystem from one of objective technician to client advocate, as they defend their role and work as developer of the repository.

Despite the DOE being given discretion in the development of a repository, the nature of conflict in the UNF subsystem meant that there were several control mechanisms within the NWPA. These included the NRC acting as a regulator of the DOE's development of a repository and the way in which the repository development was funded. I expected that the minority coalition, in a high conflict subsystem, would prefer bureaucracies that are objective technicians and given little discretion. As noted, the NRC seems to play the role of objective technician within the subsystem, as the minority coalition would prefer. In addition, the NRC role was effectively one of oversight of the DOE, which has the effect of limiting DOE's discretion. The most powerful control mechanism over DOE however, was the way which OCRWM (the agency within DOE charged with repository development) was funded. The NWPA established the Nuclear Waste Trust Fund, which collected fees from energy companies (these fees were passed to the consumers) that would be set aside by the government for the development of a permanent UNF repository. However, the DOE's budget, including the budget for OCRWM, was determined by Congressional vote within the Energy and Water Appropriations bill for each fiscal year. The implication was that despite having dedicated funds set aside for the UNF repository, budgetary decisions and allocation were still made through the political process each year. This allowed political actors to exercise substantial control over OCRWM's budget and, as a result, how it developed the UNF repository. 

The NWPA settled the question of UNF storage on deep geologic storage, however it caused conflict to increase as questions arose about the location of a potential repository. As conflict increased, so did the level of political control, particularly over the DOE. First, in the NWPAA Congress limited site characterization to Yucca Mountain. Interestingly, the choice to limit consideration was largely made by the dominant coalition, particularly Senator J. Bennett Johnston as a way to limit conflict within the subsystem. Indeed, Senator Johnston, following the passage of the NWPAA, was quoted as saying "It's fair to say we've solved the nuclear waste problem with this legislation" [@rasky_accord_1987]. Despite Senator Johnston's pronouncement, conflict in the subsystem seemed only to grow. A second way that political control over DOE increased was the creation of a new independent agency, the Nuclear Waste Technical Review Board, whose mission is to provide scientific and technical oversight of the DOE's UNF repository development. This agency, the result of a comprise between the dominant and minority coalition, was the type of objective technician agency that the minority coalition would prefer. Next, in the EnPA of 1992 Congress increased the responsibility of the EPA and their involvement in regulating radiation exposure standards at Yucca Mountain. Finally, the budget for OCRWM, designed to be decided by Congress, was consistently a point of controversy between supporters and opponents of Yucca Mountain.

The alignment of agencies within and across the Yucca Mountain coalitions were determined by the implementation structures within the NWPA, the NWPAA, and EnPA. Proposition 3 stated that the dominant and minority coalitions within a subsystem will have different preferences about the type of delegation and the role played by agencies within a subsystem, depending on the nature of the conflict within the subsystem. For the most part these expectations were met, however policy designs in conflictual subsystems are typically the result of comprise between the dominant and minority coalition. Therefore, the dominant coalition was able to ensure that the DOE, charged with the development of the repository, remained in the dominant coalition. However, the minority coalition was also able to extract compromise and limit the discretion of the DOE. 

The governance arraignments that result from policy designs, also present opportunities, in the form of opportunity structures, for the minority coalition to influence policy development [@gupta_comparative_2014]. The funding mechanism for OCRWM provided opportunities for the minority coalition to exercise control over the development of the UNF repository. The next section examines the theoretical propositions in the context of the budget for the OCRWM. 

## The OCRWM Budget and Political Control 

Political principals and bureaucratic agencies are both members of advocacy coalitions and this gives principals in competing coalitions opportunities to impact the agencies within the opposing coalition. As noted, the minority coalition in high conflict subsystems, like UNF management, would prefer to limit the discretion of agencies. Due to the substitution effect, these principals are indifferent as to whether they are able to influence agencies _ex ante_ or _ex post_. 

Budgets are one of the key ways in which politicians exert _ex post_ control over public agencies and the OCRWM's budget provided opportunities for those in the anti-Yucca coalition to limit the DOE's discretion. Figure \ref{fig:perCHG} illustrates the percent change in year-to-year budget appropriations for OCRWM from 1983-2010.^[The budget data can be found here, \url{http://www.thenwsc.org}.] The panel on the left illustrates the overall nature of budget changes, with a normal distribution overlaid, and the panel on the right shows the percent change from fiscal year-to-year. 

```{r, include=FALSE}
pdf("perCHG.pdf")
hist(bud$perCHG, prob=TRUE, xlab=("Percent Change in Budget"), main="")
curve(dnorm(x, mean=mean(na.omit(bud$perCHG)),sd=sd(na.omit(bud$perCHG))), add=TRUE, lwd=2)
dev.off()
pdf("perCHGline.pdf")
plot(bud$FY, bud$perCHG, type="l", ylim=c(-50,50), xlab=("Fiscal Year"), ylab=c("Percent Change in Budget"))
grid()
abline(h=0, lty=5)
dev.off()
```

\begin{figure}[!htp]
\centering
\caption{Percent Change Year to Year in OCRWM Budget: FY 1983-2010}
\label{fig:perCHG}
\begin{subfigure}[b]{0.4\textwidth}
\includegraphics[width=\textwidth]{perCHG.pdf}
\end{subfigure}
\begin{subfigure}[b]{0.4\textwidth}
\includegraphics[width=\textwidth]{perCHGline.pdf}
\end{subfigure}
\end{figure}


As can be seen, OCRWM did not seem to experience the familiar punctuated, or leptokurtic, pattern of budgets [@jones_politics_2005], instead the distribution seems close to normal. This is consistent with other work that has shown variation in the distribution of budget changes across policy areas [@breunig_stability_2010]. Overall and across all fiscal years, OCRWM experienced an average percent change of `r round(mean(bud$perCHG, na.rm=TRUE),4)`, which is close to 1, which would indicate and average yearly budget increase of nearly 1\%. The largest percent negative change, indicating a budget cut from the previous year was `r round(min(bud$perCHG, na.rm=TRUE),3)`\%, which occurred from FY 2009 to 2010 after the Yucca Mountain construction license that was submitted to the NRC was withdrawn. The largest percentage increase was `r round(max(bud$perCHG, na.rm=TRUE),3)`\%, from FY 1985 to 1986, following passage of the NWPA but prior to the NWPAA. Also note that following the selection of Yucca Mountain and the increased level of conflict, the OCRWM's budget was cut by `r round(bud$perCHG[[5]],3)`\% from FY 1986 to 1987 and by `r round(bud$perCHG[[6]],3)`\% from FY 1987 to 1988. 

```{r, include=FALSE}
Rbud <- tapply(bud$perCHG, bud$reid, mean, na.rm=TRUE)
summary(aovR <- aov(bud$perCHG~as.factor(bud$reid)))
aovRp <- summary(aovR)[[1]][["Pr(>F)"]]
```

The mean percent change in OCRWM's budget was, overall, slightly positive, however there was wide variation from fiscal year-to-year. As noted, the nature of the political actors within the coalitions can impact the ability of the coalition to influence policy. Powerful political principals, such as presidents or members of Congress can be an important component in coalition success. For example, Senator Johnston, as a member of the dominant coalition was paramount in the passage of the NWPA and in the selection of Yucca Mountain. In addition, President George W. Bush, also in the dominant coalition, approved Yucca Mountain as the repository site. However, the minority coalition can have a large impact on policy through having just a few powerful political actors in the coalition. For example, President Obama was able to effectively shut down the development of the Yucca Mountain repository. Prior to that however, Senator Reid from Nevada, a member of the minority anti-Yucca coalition, became the minority Senate leader in 2005. Due to his role in Senate leadership, Senator Reid was able to exercise substantial political control over the OCRWM's budget. Note that, the mean percent change in budget for the fiscal years 1984 to 2004 was `r round(Rbud[[1]],3)`\%, however the mean budget change after Senator Reid became minority leader was `r round(Rbud[[2]],4)`\%. This difference is significant at $F <$ `r round(aovRp[[1]],4)`. This illustrates the importance of which principals are in which coalition, how minority coalitions are able to use control mechanisms (e.g., agency budgets) to impact policy development, and how powerful political actors are important to advocacy coalitions. 

# Discussion and Conclusion 

The growth of the administrative state in the United States has raised questions about the ways in which that administrative state can be held accountable to elected officials and citizens. The administrative state has grown, in part, in response to the increasingly complex problems that governments face. Agency theory has been the primary tool that scholars have used to examine the ways in which elected officials allow bureaucracies some delegated policy authority. This delegation of authority is coupled with control mechanisms that allow elected official to "reign in" bureaucracies if their actions become misaligned from the elected official's preferences. The literature on agency theory has developed largely independent from the literature on the policymaking process. In an attempt to leverage these literatures to better understand the dynamics of principal-agent relationships and their impact on policymaking, I developed three theoretical propositions based on agency theory and the ACF.

Drawing from the ACF, I argued that elected officials--the principals--and bureaucrats--the agents--are both members of advocacy coalitions. This insight can contribute to the ACF through the recognition that powerful elected officials are part of advocacy coalitions. Which, while recognized, has been largely underdeveloped in the ACF literature. In addition, the ACF posits that policy actors, including bureaucrats, are policy-motivated and this motivation is driven by deep core and policy core beliefs. This insight relates to a growing body of work that addresses the values of public managers and bureaucrats and how those values impact behavior. It was evidenced by the case study of UNF management that elected officials, such as presidents and members of Congress, and executive agencies, such as the DOE, are indeed members of advocacy coalitions. Based on the ally principle, it was expected that elected officials would delegate more policymaking authority to agencies within their coalition. In the case of UNF management, this would mean that the DOE, as part of the dominant coalition, would be given rather wide discretion to develop a UNF repository. However the evidence was mixed, with the DOE seeming to have been granted discretion in the initial development of the UNF subsystem, yet this coupled with control mechanisms, like control of OCRWM's budget, that were successfully exploited to limit DOE's discretion by the minority coalition. 

The second theoretical proposition stated that multiple principals can exert influence over bureaucracies and these principals are aggregated across multiple coalitions. In the case of UNF management, the pro-Yucca coalition, over time, included the DOE, members of Congress, and President George W. Bush. While this coalition was largely dominant, the minority coalition that was opposed to Yucca Mountain included the state of Nevada, environmental groups, and powerful political actors such as Senator Reid and Presidents Clinton and Obama. With these powerful actors, the minority coalition was able to affect the policy development of Yucca Mountain. Most notably, Senator Reid was successful in cutting funding for the development of Yucca Mountain. 

Finally, I posited that delegation and control mechanisms are embedded within policy designs and that elected officials would prefer different roles and different levels of delegation based on the type of coalition they are in, dominant or minority, and the level of conflict in the subsystem, low or high. It is expected that the dominant coalition always prefers to grant discretion, yet prefers objective technicians when conflict is low and client advocates when conflict is high. The minority coalition is expected to prefer client advocates with high discretion in subsystems with low conflict and objective technicians with low discretion in subsystems that are high in conflict. However, policy designs are often created as compromises, therefore it is likely that there would be a mix of types and levels of discretion. This seemed to be the case with UNF management, given that the amount of discretion given to the DOE to develop a repository was limited through policy change as the UNF subsystem development and became increasing conflictual. Some of the ways in which the DOE's discretion was limited included making other agencies such as the NRC, the NWTRB, and the EPA regulators of the DOE with regard to the development of the UNF repository. The nature of the implementation structure within UNF policies, illustrates how policy designs, and as a result, the nature of delegation and control are often the products of compromise between advocacy coalitions. 

This manuscript sketches some preliminary ideas about how agency theory and the ACF can be combined to better understand the nature of shared policymaking authority between elected officials and bureaucrats that exist within and across advocacy coalitions. Using the case study of UNF management, I found tentative support for a set of conjectures about the nature and delegation and control within a contentious subsystem, yet much work needs to be done. First, better measures of delegation and/or discretion need to be developed. Using case studies and policy histories can be very effective in building insights about delegation, however measures of delegation that could be applied across a vary of policy areas and subsystem types would be more effective for theory building  and hypothesis testing. The Institutional Grammar Tool could, perhaps, be adapted for this purpose [@basurto_systematic_2010: @siddiki_dissecting_2011].

A second necessary advancement entails better incorporating notions of strategic agents within agency theory and the ACF. The propositions that were presented here are largely top-down and focused on the elected officials as the principals. More work should focus on how bureaucrats, as policy-motivated members of advocacy coalitions leverage there involvement in coalitions to achieve their goals. 



\newpage

# References
\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{8pt}
\vspace*{-0.2in}
\singlespace
\noindent

